\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-03-01 submission/manuscript.Rmd

---
title: "**Evaluation of binary classification pipelines and methods for 16S rRNA gene data**"
bibliography: references.bib
output:
  pdf_document: 
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

@@

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration and taxonomic characterization of the human associated microbiome. Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health. However, this is an undertaking as human associated microbial communities are remarkably uneven. It is unlikely that a single species can explain a disease state comprehensively. It is more likely that subsets of those communities, in relation to one another and to their host, account for different disease states. Thus, researchers have started to explore the utility of machine learning (ML) techniques to develop predictive models that use microbiota associated biomarkers to identify healthy and diseased states. However, currently the field's use of ML lacks clarity and consistency on which methods are used, how these methods are implemented and if these methods are robust and reproducible. Moreover, there is a lack of consideration on why a particular ML model is utilized. There is a wide range of complexity and interpretibility in models used in microbiome field [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @pasolli_machine_2016]; recently with many that are black boxes which cannot be interpreted by humans [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017]. ML model choice should reflect the goal of the study; whether it is to develop the best predictive model or to understand the ecology of microbiota associated diseases.

One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers. Colorectal cancer (CRC) is one of the leading causes of death in the US, therefore developing a predictive model from a stool microbiome assay as a non-invasive screening tool is a remarkable innovation. However, the efforts to combine ML methods with relative abundances of bacterial poplations in stool to predict CRC tumors have been afflicted by flawed ML methods and by the lack of discussion on modeling choices. In previous studies, ... There is a need for implementing consistent and transparent machine learning practices to generate reproducible and replicable CRC biomarker models. This study aims to shed light on how much task definition and model choices can affect the results. Here we performed an empirical analysis comparing several different modeling pipelines using the same dataset.

As our dataset, we used a previously published CRC study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and immunochemical test (FIT) results from `r paste(nrow(shared))` patients. We built ML models using 16S abundances and FIT to predict patients with normal or screen relevant neoplasias (SRN) disease status. The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. These models increase in complexity while they decrease in interpretability. Our ML pipeline performed 100 data-splits and utilized held-out test data to evaluate generalization and predicton performance of each ML model. The mean AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`). Random Forest and XGBoost had the highest mean AUROC for detecting SRN. In terms of computational efficiency, L2 Logistic Regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[6],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`). We found that mean cross-validation and testing AUROC could vary by as much as `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set for evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

>>>>

---
title: "**Evaluation of machine learning methods for 16S rRNA gene data**"
bibliography: references.bib
output:
  pdf_document: 
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

@@

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration and taxonomic characterization of the human associated microbiome. Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for reproducible methods to identify associations between members of the microbiome and human host. However, this is an undertaking as human associated microbial communities are remarkably uneven. It is unlikely that a single species can explain a phenotype. Instead, subsets of those communities, in relation to one another and to their host, account for the differences. Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict a phenotype and to understand the microbial ecology of the host. However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented. Moreover, there is a lack of consideration on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017] without a discussion on if interpretibility is necessary for the study. If we want to generate reproducible and robust microbiota-associated biomarker models we need to (1) implement consistent, accessible and transparent machine learning practices; (2) use ML models that reflect the goal of the study as it will inform/define the expectations of model accuracy, complexity, interpretibility, computational efficiency and scalability.

One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers in the stool [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017]. Colorectal cancer (CRC) is one of the leading causes of death in the US, therefore developing a non-invasive screening tool with ML is a remarkable innovation. However, afforementioned standards have not been established in these ML models. Here we performed an empirical analysis comparing several different modeling pipelines using the same dataset. We used a previously published CRC study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and immunochemical test (FIT) results from `r paste(nrow(shared))` patients. We built ML models using 16S abundances and FIT to predict patients with normal or screen relevant neoplasias (SRN) disease status. The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. These models increase in complexity while they decrease in interpretability. Our ML pipeline performed 100 data-splits and utilized held-out test data to evaluate generalization and predicton performance of each ML model. The mean AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`). Random Forest and XGBoost had the highest mean AUROC for detecting SRN. In terms of computational efficiency, L2 Logistic Regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[6],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`). We found that mean cross-validation and testing AUROC could vary by as much as `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set for evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

### 2019-03-01 submission/manuscript.tex

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Evaluation of binary classification pipelines and methods for 16S rRNA gene data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Evaluation of binary classification pipelines and methods for
16S rRNA gene data}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration and taxonomic
characterization of the human associated microbiome. Currently, the
microbiome field is growing at an unprecedented rate and as a result,
there is an ever-increasing demand for reproducible methods for
identifying associations between members of the microbiome and human
health. However, this is an undertaking as human associated microbial
communities are remarkably uneven. It is unlikely that a single species
can explain a disease state comprehensively. It is more likely that
subsets of those communities, in relation to one another and to their
host, account for different disease states. Thus, researchers have
started to explore the utility of machine learning (ML) techniques to
develop predictive models that use microbiota associated biomarkers to
identify healthy and diseased states. However, currently the field's use
of ML lacks clarity and consistency on which methods are used, how these
methods are implemented and if these methods are robust and
reproducible. Moreover, there is a lack of consideration on why a
particular ML model is utilized. There is a wide range of complexity and
interpretibility in models used in microbiome field (1--5); recently
with many that are black boxes which cannot be interpreted by humans
(6--8). ML model choice should reflect the goal of the study; whether it
is to develop the best predictive model or to understand the ecology of
microbiota associated diseases.

One application of machine learning to microbiome data has been to
classify patients as having colorectal tumors based on
microbiota-associated biomarkers. Colorectal cancer (CRC) is one of the
leading causes of death in the US, therefore developing a predictive
model from a stool microbiome assay as a non-invasive screening tool is
a remarkable innovation. However, the efforts to combine ML methods with
relative abundances of bacterial poplations in stool to predict CRC
tumors have been afflicted by flawed ML methods and by the lack of
discussion on modeling choices. In previous studies, \ldots{} There is a
need for implementing consistent and transparent machine learning
practices to generate reproducible and replicable CRC biomarker models.
This study aims to shed light on how much task definition and model
choices can affect the results. Here we performed an empirical analysis
comparing several different modeling pipelines using the same dataset.

As our dataset, we used a previously published CRC study (3) which had
fecal 16S rRNA gene sequences and immunochemical test (FIT) results from
490 patients. We built ML models using 16S abundances and FIT to predict
patients with normal or screen relevant neoplasias (SRN) disease status.
The study had 261 normal and 229 SRN samples. We established modeling
pipelines for L2-regularized logistic regression, L1 and L2 support
vector machines (SVM) with linear and radial basis function kernels, a
decision tree, random forest and XGBoost. These models increase in
complexity while they decrease in interpretability. Our ML pipeline
performed 100 data-splits and utilized held-out test data to evaluate
generalization and predicton performance of each ML model. The mean
AUROC varied from 0.67 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest
and XGBoost had the highest mean AUROC for detecting SRN. In terms of
computational efficiency, L2 Logistic Regression trained the fastest
(0.098 hours, std ± 0.008), while XGBoost took the longest (3.601 hours,
std ± 0.524). We found that mean cross-validation and testing AUROC
could vary by as much as 0.062, which highlights the importance of a
seperate held-out test set for evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data using AUROC as our metric.
Random Forest and XGBoost had the highest mean AUROC for detecting SRN,
0.818 (std ± 0.036) and 0.814 (std ± 0.038) respectively {[}Figure 2{]}.
L1 linear SVM and decision tree had significantly lower AUROC values,
0.748 (std ± 0.043) and 0.741 (std ± 0.038) {[}Figure 2{]}. However,
they had significanly higher performances than L2 linear SVM, RBF SVM
and L2 logistic regression which had mean AUROC values of 0.67 (std ±
0.054), 0.683 (std ± 0.048) and 0.677 (std ± 0.053)
respectively{[}Figure 2{]}. We also evaluated the generalization
performance of each classifier by comparing their mean cross-validation
AUROC and mean testing AUROC. We found a significant difference for
classifiers L2 support vector machines (SVM) with linear and radial
basis function kernels and L2 logistic regression. These differences
were 0.062, 0.047 and 0.061, respectively {[}Figure 2{]} .

\textbf{The prediction performance of classifiers for different
hyperparameter settings.}

@@

\hypertarget{refs}{}
\hypertarget{ref-zeller_potential_2014}{}
1. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Mol Syst Biol
\textbf{10}.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\hypertarget{ref-zackular_human_2014}{}
2. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.

\hypertarget{ref-baxter_dna_2016}{}
3. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
immunochemical test can replace stool for detection of colonic lesions
using a microbiota-based model. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.

\hypertarget{ref-baxter_microbiota-based_2016}{}
4. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-pasolli_machine_2016}{}
5. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
meta-analysis of large metagenomic datasets: Tools and biological
insights. PLoS Comput Biol \textbf{12}.
doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.

\hypertarget{ref-galkin_human_2018}{}
6. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
2018. Human microbiome aging clocks based on deep learning and tandem of
permutation feature importance and accumulated local effects. bioRxiv.
doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.

\hypertarget{ref-reiman_using_2017}{}
7. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
convolutional neural networks to explore the microbiome, pp. 4269--4272.
\emph{In} 2017 39th annual international conference of the IEEE
engineering in medicine and biology society (EMBC).

\hypertarget{ref-fioravanti_phylogenetic_2017}{}
8. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Evaluation of machine learning methods for 16S rRNA gene data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Evaluation of machine learning methods for 16S rRNA gene data}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration and taxonomic
characterization of the human associated microbiome. Currently, the
microbiome field is growing at an unprecedented rate and as a result,
there is an increasing demand for reproducible methods to identify
associations between members of the microbiome and human host. However,
this is an undertaking as human associated microbial communities are
remarkably uneven. It is unlikely that a single species can explain a
phenotype. Instead, subsets of those communities, in relation to one
another and to their host, account for the differences. Therefore,
researchers have started to explore the utility of machine learning (ML)
models that use microbiota associated biomarkers to predict a phenotype
and to understand the microbial ecology of the host. However, currently
the field's use of ML lacks clarity and consistency on which methods are
used and how these methods are implemented. Moreover, there is a lack of
consideration on why a particular ML model is utilized. Recently, there
is a trend towards using more complex ML models (1--3) without a
discussion on if interpretibility is important for the study. If we want
to generate reproducible and robust microbiota-associated biomarker
models we need to (1) implement consistent, accessible and transparent
machine learning practices; (2) use ML models that reflect the goal of
the study as it will inform/define the expectations of model accuracy,
complexity, interpretibility, computational efficiency and scalability.

One application of machine learning to microbiome data has been to
classify patients as having colorectal tumors based on
microbiota-associated biomarkers in the stool (4--8). Colorectal cancer
(CRC) is one of the leading causes of death in the US, therefore
developing a non-invasive screening tool with ML is a remarkable
innovation. However, afforementioned standards have not been established
in these ML models. Here we performed an empirical analysis comparing
several different modeling pipelines using the same dataset. We used a
previously published CRC study (6) which had fecal 16S rRNA gene
sequences and immunochemical test (FIT) results from 490 patients. We
built ML models using 16S abundances and FIT to predict patients with
normal or screen relevant neoplasias (SRN) disease status. The study had
261 normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. These models increase in complexity while
they decrease in interpretability. Our ML pipeline performed 100
data-splits and utilized held-out test data to evaluate generalization
and predicton performance of each ML model. The mean AUROC varied from
0.67 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest and XGBoost had
the highest mean AUROC for detecting SRN. In terms of computational
efficiency, L2 Logistic Regression trained the fastest (0.098 hours, std
± 0.008), while XGBoost took the longest (82.47 hours, std ± 17.045). We
found that mean cross-validation and testing AUROC could vary by as much
as 0.062, which highlights the importance of a seperate held-out test
set for evaluation. Aside from evaluating generalization and
classification performances for each of these models, this study
established standards for modeling pipelines of microbiome-associated
machine learning models.

@@

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data using AUROC as our metric.
Random Forest and XGBoost had the highest mean AUROC for detecting SRN,
0.816 (std ± 0.039) and 0.814 (std ± 0.038) respectively {[}Figure 2{]}.
L1 linear SVM and decision tree had significantly lower AUROC values,
0.748 (std ± 0.043) and 0.741 (std ± 0.038) {[}Figure 2{]}. However,
they had significanly higher performances than L2 linear SVM, RBF SVM
and L2 logistic regression which had mean AUROC values of 0.67 (std ±
0.054), 0.684 (std ± 0.048) and 0.677 (std ± 0.053) respectively
{[}Figure 2{]}. We also evaluated the generalization performance of each
classifier by comparing their mean cross-validation AUROC and mean
testing AUROC. We found a statistically significant difference for
classifiers L2 support vector machines (SVM) with linear and radial
basis function kernels and L2 logistic regression. These differences
were 0.062, 0.046 and 0.061, respectively {[}Figure 2{]} .

\textbf{Hyperparameter optimization for each classifier.}

@@

\hypertarget{refs}{}
\hypertarget{ref-galkin_human_2018}{}
1. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
2018. Human microbiome aging clocks based on deep learning and tandem of
permutation feature importance and accumulated local effects. bioRxiv.
doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.

\hypertarget{ref-reiman_using_2017}{}
2. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
convolutional neural networks to explore the microbiome, pp. 4269--4272.
\emph{In} 2017 39th annual international conference of the IEEE
engineering in medicine and biology society (EMBC).

\hypertarget{ref-fioravanti_phylogenetic_2017}{}
3. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.

\hypertarget{ref-zeller_potential_2014}{}
4. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Mol Syst Biol
\textbf{10}.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\hypertarget{ref-zackular_human_2014}{}
5. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.

\hypertarget{ref-baxter_dna_2016}{}
6. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
immunochemical test can replace stool for detection of colonic lesions
using a microbiota-based model. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.

\hypertarget{ref-baxter_microbiota-based_2016}{}
7. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-hale_shifts_2017}{}
8. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the fecal
microbiota associated with adenomatous polyps. Cancer Epidemiol
Biomarkers Prev \textbf{26}:85--94.
doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.
